---
title: "Statistics in R with the `tidyverse`"
subtitle: "Day 1 Walkthrough"
author: "Dr. Chester Ismay"
format: html
---

```{r}
#| include: false
options(width = 120)
```


# Working with Data in R - Explore, Visualize, Wrangle, Import

## Session 1: Introduction to R and RStudio

### 1. Installing R and RStudio

- You need to install R first from <https://cloud.r-project.org/> and then install RStudio from <https://posit.co/download/rstudio-desktop/>. 
- Once installed, work in RStudio to interact with R efficiently.

---

### 2. Exploring the RStudio Interface

- In RStudio, you will see three panes: Console, Environment, and Files.
- The Console is where you type and run your R code.
- The Environment pane shows all objects (like datasets) currently in memory.
- The Files pane helps you navigate files in your project.

---

### 3. Installing Packages

```{r}
#| eval: false
# To install a package, use the install.packages() function.

# or together using the c() function:

```


```{r}
#| eval: false

# Install all necessary packages

```


---

### 4. Loading Packages

```{r}
#| message: false
# To load a package, use the library() function.


# TIP: Check help with `?` operator or `help()` function.

```

---

### 5. Viewing a Dataset

```{r}
#| eval: false

# Can load the data into the Environment pane


# To view a dataset in RStudio, use the View() function

```

- Identification variables: 
- Measurement variables: 

---

### 6. Exploring Dataset Structure and Data Types

```{r}
# To see an overview of the dataset’s structure:

```

---

### 7. Accessing a Single Column

```{r}
# To access a specific column, use the `$` operator:

```

---

### 8. Checking the First Few Rows

```{r}
# To quickly see the first 6 rows of the dataset:

```

---

### 9. Basic Operations in R

```{r}
# You can do basic math in R, like addition:


# Or create sequences:

```

---

### 10. Using Functions in R

```{r}
# Using named arguments in functions:


# Or using positional arguments:

```

---

### Session 1 Review Questions

**(1.1)** Which of the following is required before you can use RStudio effectively?

A. You need to install RStudio first, as R is automatically included.  
B. R should be installed from the R Project website, and then RStudio can be installed separately.  
C. R is pre-installed on most operating systems, so only RStudio needs to be downloaded.  
D. R and RStudio can be used without any installation, directly from a web browser.

---

**(1.2)** What is the purpose of the Environment pane in RStudio?

A. To display the output of the R code you run, including plots and messages.  
B. To show all objects and datasets currently loaded in your R session.  
C. To provide a workspace for writing R scripts and functions.  
D. To navigate and organize the files within your project folder.

---

**(1.3)** How do you install multiple R packages in one step?

A. Use the `install.packages()` function with each package listed individually: `install.packages("ggplot2"), install.packages("dplyr")`.  
B. You must install each package one by one, as R doesn't allow installing multiple packages in a single command.  
C. Use the `install.packages()` function with the `c()` function to install several packages at once.  
D. You can install multiple packages by using the `load()` function for each package separately.

---

**(1.4)** Which of the following correctly loads the `dplyr` package into your R session?

A. `load(dplyr)`  
B. `library(dplyr)`  
C. `install.packages("dplyr")`  
D. `View(dplyr)`

---

**(1.5)** What does the `glimpse()` function from the `dplyr` package do?

A. It provides a detailed report of summary statistics for each variable in the dataset.  
B. It opens the dataset in a spreadsheet-like viewer for easy exploration.  
C. It gives a quick overview of the structure of a dataset, showing the data types and the first few entries of each column.  
D. It creates a visual summary of numeric columns using histograms and scatterplots.

---

## Session 2: Data Visualization with `ggplot2`

### 11. Installing and Loading Necessary Packages

```{r}
# Load required packages for data visualization


# TIP: You can install packages if not already installed

```

---

### 12. Visualizing Population Distribution: Histogram

```{r}
# Create a histogram of population distribution

```

---

### 13. Visualizing Life Expectancy: Side-by-side Boxplot

```{r}
# Create a boxplot to see GDP per capita spread by continent

```


---

### 14. Comparing Number of Countries by Continent: Barplot

```{r}
# Create a barplot to compare the number of countries by continent

```

---

### 15. Scatterplot: GDP per Capita vs. Life Expectancy

```{r}
# Create a scatterplot to show the relationship between GDP per capita and life expectancy

```

Check out [Subsection 2.3.2 of ModernDive V2](https://moderndive.com/v2/viz.html#overplotting) for overplotting discussion.

---

### 16. Faceted Scatterplot: GDP per Capita vs. Life Expectancy by Continent

```{r}
# Faceted scatterplot to see relationship by continent

```


---

### 17. Pie Chart vs Bar Chart: Region Distribution

```{r}
# Create a pie chart for region distribution

```


```{r}
# As a bar chart instead

```

---

### 18. Line Graph: Time Series Data


```{r}
#install.packages("fivethirtyeight")
library(fivethirtyeight)

# Create a line graph showing the number of births
```


```{r}
#| message: false
# Include package for more easily working with dates
# install.packages("lubridate")
library(lubridate)
library(dplyr)

# Filter the data for 2014
US_births_2014 <- US_births_2000_2014 |>
  filter(year(date) == 2014)

# Create a line graph showing the number of births in 2014

```

### Session 2 Review Questions

**(2.1)** In the following code for creating a scatterplot of Fertility Rate (`fertility_rate_2022`) vs. Human Development Index (`hdi_2022`), which entries should be in quotes?

```r
ggplot(un_member_states_2024, aes(x = fertility_rate_2022, y = hdi_2022)) + 
  geom_point()
```

A. `un_member_states_2024`, `fertility_rate_2022`, and `hdi_2022` should all be in quotes.  
B. Only `un_member_states_2024` should be in quotes, as it refers to the dataset.  
C. Only `fertility_rate_2022` and `hdi_2022` should be in quotes, as they are variable names.  
D. None of the elements in the code should be in quotes.

---

**(2.2)** How can you create a histogram of the population distribution in the `un_member_states_2024` data frame?

A. `geom_bar(aes(x = population_2024))`  
B. `ggplot(aes(y = population_2024)) + geom_boxplot()`  
C. `ggplot(un_member_states_2024, aes(x = population_2024)) + geom_histogram()`  
D. `ggplot(un_member_states_2024, aes(x = continent)) + geom_bar()`  

---

**(2.3)** What does a boxplot created using the `ggplot()` function reveal about life expectancy?

A. The boxplot reveals the mean and standard deviation of life expectancy values.  
B. It shows how life expectancy is spread and identifies outliers.  
C. A boxplot represents the frequency of different life expectancy ranges.  
D. It is used to display the total sum of life expectancy values for each continent.  

---

**(2.4)** How can you compare the number of countries by continent using a barplot?

A. `ggplot(un_member_states_2024, aes(x = continent)) + geom_bar()`  
B. `ggplot(un_member_states_2024, aes(x = gdp_per_capita)) + geom_bar()`  
C. `ggplot(un_member_states_2024, aes(x = continent, y = life_expectancy_2022)) + geom_bar()`  
D. `ggplot(un_member_states_2024, aes(x = continent, fill = population_2024)) + geom_bar()`  

---

**(2.5)** What is an appropriate purpose of facet wrapping in a scatterplot showing Capital Population vs. Obesity Rate 2024?

A. To split the scatterplot into multiple graphs by region.  
B. To add color to each region based on obesity rate.  
C. To combine the population of the capital city and the country's obesity rate in a single graph.  
D. To display the relationship between capital city population and obesity rate over time.  

---

## Session 3: Data Wrangling and Tidy Data

### 19. Filtering Rows with `filter()`

```{r filter-example}
# Filter countries in Africa with GDP per capita greater than $5,000


# TIP: Can also use the `&` instead of `,` for AND conditions


# This is not the same as using `|` which is an OR condition

```

---

### 20. Summarizing Data with `summarize()`

```{r summarize-example}
# Summarize average life expectancy and population

# Can also display as a markdown table
library(knitr)

```

---

### 21. Grouping Data with `group_by()`

```{r groupby-example}
# Group by continent and summarize the average life expectancy for each group

```

---

### 22. Creating New Variables with `mutate()`

```{r mutate-example}
# Create a new variable that categorizes countries by GDP per capita

```

---

### 23. Arranging Rows with `arrange()`

```{r arrange-example}
# Arrange countries by population in descending order

```

---

### 24. Selecting Specific Columns with `select()`

```{r select-example}
# Select country name, continent, and population

```

---

### 25. Putting a Whole `|>`line Together

```{r}
# Create a pipeline that filters, groups, summarizes, mutates, arranges, 
# and selects columns for countries in Asia and Europe

```

### 26. Creating data for `geom_col()` use

```{r}
# Create a data frame for geom_col() use to count number of countries per region

```

### 27. Converting wide data into tidy data

```{r}
# Convert wide data into tidy data
library(tibble)
library(tidyr)

# Collected from https://data.worldbank.org/?locations=BR-NG-ID
wide_unemp <- tibble(
  country = c("Brazil", "Nigeria", "Indonesia"),
  `2021` = c(13.2, 5.4, 3.8),
  `2022` = c(9.2, 3.8, 3.5),
  `2023` = c(8, 3.1, 3.4)
)

# Use pivot_longer to convert data


# Plot data as a linegraph

```

### 28. Importing data from a CSV file

```{r}
library(readr)

# Load data from data_dev_survey.csv in same file path as this .qmd file

```

---

### Session 3 Review Questions

**`(3.1)`** What is the role of the pipe operator `|>` in streamlining data wrangling tasks?

A. It combines two datasets into one, merging data from different sources into a unified data frame.  
B. It allows the output of one function to be passed to the next, improving readability and making code writing more efficient.   
C. It saves a dataset to a file, preserving changes by writing the transformed data to a CSV or another format.  
D. It generates plots automatically from a dataset, creating graphs without needing to specify the plotting functions. 

**`(3.2)`** How would you filter countries in the `un_member_states_2024` data frame that are in Asia and have a population greater than 100 million?

A. `filter(un_member_states_2024, continent = "Asia" & population_2024 > 1e8)`  
B. `un_member_states_2024 |> select(continent = "Asia" & population_2024 > 100000000)`  
C. `un_member_states_2024 |> filter(continent == "Asia", population_2024 > 100000000)`  
D. `un_member_states_2024 |> filter(population_2024 > 100000000 | continent == "Asia")`  

**`(3.3)`** Using `group_by()` and `summarize()`, how can you calculate the total population for each continent where the line of code directly above is `us_member_states_2024 |>`?

A. `group_by(continent) |> summarize(total_population = mean(population_2024))`  
B. `group_by(continent) |> summarize(total_population = sum(population_2024))`  
C. `summarize(continent, total_population = sum(population_2024))`  
D. `group_by(population_2024) |> summarize(total_population = sum(continent))`  

**`(3.4)`** How does the `mutate()` function help in creating new categorical variables?

A. `mutate()` always replaces an existing variable with a new one based on a conditional statement.  
B. `mutate()` drops variables from a dataset based on specific conditions.  
C. `mutate()` filters rows by removing missing values from a variable.  
D. `mutate()` adds a new variable based on the transformation of existing variables.  

**`(3.5)`** How can you arrange countries by their GDP per capita in ascending order?

A. `arrange(un_member_states_2024, gdp_per_capita)`    
B. `arrange(un_member_states_2024 |> gdp_per_capita)`  
C. `group_by(gdp_per_capita) |> arrange(continent)`  
D. `arrange(gdp_per_capita, un_member_states_2024)`  
