---
title: "Statistics in R with the `tidyverse`"
author: "Dr. Chester Ismay"
subtitle: "Day 2 Walkthrough"
format: html
---

```{r}
#| include: false
options(width = 120)
```


# Day 2: Linear Regression in R

## Session 1: Simple Linear Regression Analysis

### 1. Load Necessary Packages

```{r}
# Load the required packages
library(dplyr)
library(ggplot2)
library(moderndive)
```

- These packages provide tools for data wrangling, visualization, and modeling.  

---

### 2. Prepare Dataset

```{r}
library(palmerpenguins)

# Only use complete rows of data
penguins_data <- penguins |> 
  na.omit()
```

- We'll be exploring the `penguins` data frame from the `palmerpenguins` package as well. It contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antarctica. More information is [here](https://allisonhorst.github.io/palmerpenguins/articles/intro.html).  
- We remove rows with missing values.

---

### 3. Exploring Dataset Structure

```{r}
# Use glimpse to explore the structure of the dataset
glimpse(penguins_data)
```


---

### 4. Summary Statistics

```{r}
# Summarize mean and median values for flipper length and body mass
penguins_data |> 
  summarize(mean_flipper_length = mean(flipper_length_mm),
            mean_body_mass = mean(body_mass_g),
            median_flipper_length = median(flipper_length_mm),
            median_body_mass = median(body_mass_g))
```


---

### 5. Using `tidy_summary()` Function

```{r}
# Use tidy_summary to get a clean summary of the data
penguins_data |> 
  select(flipper_length_mm, body_mass_g) |> 
  tidy_summary()
```


---

### 6. Correlation Between Flipper Length and Body Mass

```{r}
# Compute the correlation between flipper length and body mass
penguins_data |> 
  get_correlation(formula = body_mass_g ~ flipper_length_mm)
```


---

### 7. Scatterplot of Flipper Length and Body Mass

```{r}
# Plot the relationship between flipper length and body mass
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.5, color = "turquoise")
```


---

### 8. Regression Line

```{r}
# The same scatterplot with a regression line added
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.5, color = "turquoise") +
  geom_smooth(method = "lm", se = FALSE)
```


---

### 9. Fit a Regression Model

```{r}
# Fit a linear regression model
flipper_model <- lm(body_mass_g ~ flipper_length_mm, data = penguins_data)

# Get the regression coefficients
coef(flipper_model)
```


---

### Session 1 Review Questions

**(1.1)** What is the purpose of the `na.omit()` function in the code below?

```r
penguins_data <- penguins |>
  select(species, island, flipper_length_mm, body_mass_g) |>
  na.omit()
```

A. It replaces missing values with the median.  
B. It removes any rows that contain missing values.  
C. It converts missing values to zeros.  
D. It fills missing values with the previous non-missing value.

---

**(1.2)** What does the `tidy_summary()` function do when applied to numeric columns in the `penguins_data` data frame?

A. It generates summary statistics for each column in the data frame.  
B. It prints the first 6 rows of the data frame.  
C. It provides a compact overview of the data, including column names and data types.  
D. It creates a scatterplot of the variables in the data frame.

---

**(1.3)** Which of the following correctly calculates the mean body mass in the `penguins_data` data frame?

A. `summarize |> penguins_data(mean_flipper = mean(body_mass_g))`  
B. `summarize(mean(body_mass_g))`  
C. `summarize(penguins_data, mean = body_mass_g)`
D. `penguins_data |> summarize(mean_body_mass = mean(body_mass_g))`  

---

**(1.4)** What does the `get_correlation()` function do when applied in the code below?

```r
penguins_data |>
  get_correlation(formula = flipper_length_mm ~ body_mass_g)
```

A. It fits a linear regression model.  
B. It gives a measure of the linear relationship between flipper length and body mass.
C. It plots a scatterplot of flipper length and body mass.  
D. It generates summary statistics for flipper length and body mass.

---

**(1.5)** In the following code, what is the purpose of `geom_smooth(method = "lm", se = FALSE)`?

```r
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Flipper Length (mm)", y = "Body Mass (g)",
       title = "Flipper Length vs. Body Mass with Regression Line")
```

A. It adds a smoothed curve based on a polynomial fit to the scatterplot.  
B. It adds a linear regression line to the scatterplot without displaying the confidence interval.  
C. It adjusts the transparency of the points in the scatterplot.  
D. It calculates and displays residuals on the plot.

---

## Session 2: Multiple Linear Regression Analysis (Part 1)

### 10: Fitting a Simple Linear Regression with a Categorical Regressor

```{r}
# Get the distinct levels of species
penguins_data |> distinct(species)

# Fit a simple linear regression model with species as a predictor
species_model <- lm(body_mass_g ~ species, data = penguins_data)

# Get regression coefficients
coef(species_model)
```


---

### 11: Visualizing Data with Categorical Variables

```{r}
# Scatterplot with species as color for grouping
ggplot(penguins_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

---

### 12: Summarizing the Target and Regressors (One numerical and one categorical)

```{r}
# Summarize the target and regressors
penguins_data |> 
  select(body_mass_g, flipper_length_mm, species) |> 
  tidy_summary()
```

---

### 13: Fitting a Multiple Regression Model with Interaction Terms

```{r}
# Fit a multiple regression model with interaction
multiple_model_interaction <- lm(body_mass_g ~ flipper_length_mm + species + flipper_length_mm * species,
                                 data = penguins_data)

# Same as
multiple_model_interaction <- lm(body_mass_g ~ flipper_length_mm * species,
                                 data = penguins_data)

# Get regression coefficients
coef(multiple_model_interaction)
```


---

### 14: Adding Interaction Terms to the Regression Plot

```{r}
# Scatterplot with regression lines by species
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```


---

### Session 2 Review Questions


**(2.1)** What is the default baseline island in the simple linear regression model for predicting `body_mass_g` using `island` as a categorical variable?

A. Torgersen  
B. Biscoe  
C. Dream  
D. The island with the highest body mass

---

**(2.2)** In the following regression model with interaction terms, what does the coefficient for `flipper_length_mm:speciesGentoo` represent?

```r
multi_model_interaction <- lm(body_mass_g ~ flipper_length_mm * species, data = penguins_data)
```

A. The slope of the regression line for Gentoo penguins  
B. The average body mass of Gentoo penguins  
C. The change in the relationship between flipper length and body mass for Gentoo penguins compared to the baseline  
D. The difference in intercept for Gentoo penguins compared to the baseline  

---

**(2.3)** What is the purpose of adding interaction terms to a multiple regression model?

A. To make the model more complex without any real benefit  
B. To estimate the relationship between each explanatory variable and the response variable independently  
C. To account for how the relationship between one explanatory variable and the response depends on another explanatory variable  
D. To automatically improve the model's R-squared value

---

**(2.4)** What does the following plot indicate about the relationship between flipper length and body mass for different islands?

```{r}
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g, color = island)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Flipper Length (mm)", y = "Body Mass (g)", color = "Island")
```

A. The relationship between flipper length and body mass is the same for all islands 
B. The relationship between flipper length and body mass differs across species, with each island having a different slope  
C. The islands do not influence the relationship between flipper length and body mass  
D. Only the intercept differs for each island, but the slope is the same

---

**(2.5)** In the multiple regression model with interaction terms, how are dummy variables used for categorical predictors?

A. Dummy variables are used to represent the different categories of a categorical variable 
B. Dummy variables are used to represent each numerical variable  
C. Dummy variables are used to replace the intercept in the regression model  
D. Dummy variables are not necessary for categorical variables

---

## Session 3: Multiple Linear Regression Analysis (Part 2)

### 15. Fitting a Multiple Regression Model Without Interactions

```{r}
# Fit a multiple regression model without interaction terms


# Get regression coefficients

```


---

### 16. Visualizing the Parallel Slopes Model

```{r}
# Scatterplot with regression lines by species (different intercepts but same slopes)

```


---

### 17. Summarize the Numerical Regressors and Target Variable

```{r}
# Use tidy_summary to get a clean summary of the data for a multiple regression
# with flipper length and bill length and regressors and body mass as the
# outcome

```


---

### 18. Fitting a Multiple Regression Model with Two Numerical Predictors

```{r}
# Fit a multiple regression model


# Get regression coefficients

```

---

### 19. Visualizing the Relationship between Variables (with SLR)

```{r}
# Create scatterplot with regression lines for both variables

```

---

### Session 3 Review Questions

**`(3.1)`** What is the main assumption in a multiple regression model without interaction terms?

A. The relationship between the explanatory variables and the response is always quadratic.  
B. The slope of the regression line is different for each level of the categorical variable.  
C. The slope of the regression line is the same for all levels of the categorical variable.  
D. The intercept of the regression line is the same for all levels of the categorical variable.

---

**`(3.2)`** In the model `lm(body_mass_g ~ flipper_length_mm + species, data = penguins_data)`, what does the coefficient for `speciesGentoo` represent?

A. The effect of being a Gentoo penguin on body mass compared to the baseline species (Adélie).  
B. The effect of flipper length on body mass for Gentoo penguins only.  
C. The change in flipper length due to species.  
D. The effect of body mass on flipper length for Gentoo penguins.

---

**`(3.3)`** What does the function `geom_parallel_slopes()` do in a regression plot?

A. It fits regression lines with different slopes for each group.  
B. It plots regression lines with the same slope but different intercepts for each group.  
C. It creates a scatterplot without any regression lines.  
D. It visualizes interaction effects between the variables.

---

**`(3.4)`** In the model `lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins_data)`, what do the coefficients for `flipper_length_mm` and `bill_length_mm` represent?

A. The total body mass of each penguin species.  
B. The predicted body mass for penguins with average bill and flipper lengths.
C. The interaction effect between flipper length and bill length.  
D. The effect of bill length on body mass, controlling for flipper length, and vice versa. 

---

**`(3.5)`** What does a regression model without interaction terms assume about the relationship between the explanatory variables and the response variable?

A. The explanatory variables are not related to the response.  
B. The relationship between the explanatory variables and the response is independent of one another.  
C. The explanatory variables interact to affect the response.  
D. The response variable depends only on the categorical variables.

