---
title: "Statistics in R with the `tidyverse`"
author: "Dr. Chester Ismay"
subtitle: "Day 3 Walkthrough"
format: html
---

```{r}
#| include: false
options(width = 120)
```


# Day 3: Sampling and Estimation in R

## Session 1: Sampling

### 1. Load Necessary Packages

```{r}
# Load the required packages
library(dplyr)
library(ggplot2)
library(moderndive)
library(infer)
```

- These packages provide tools for data wrangling, visualization, and modeling.  

---

### 2. Prepare Population Dataset with Generation

```{r}
# For reproducibility
set.seed(2024)

# Create vectors of sport ball types and their proportions
types_of_sport_balls <- c("Basketball", "Pickleball", "Tennis Ball", 
                          "Football/Soccer", "American Football")
proportion_of_each_type <- c(0.2, 0.15, 0.3, 0.25, 0.1)

# Create a tibble of 1200 sport balls that will act as our population
store_ball_inventory <- tibble(
  ball_ID = 1:1200,
  ball_type = sample(x = types_of_sport_balls,
                     size = 1200, 
                     replace = TRUE, 
                     prob = proportion_of_each_type
  )
)
```

- We'll be exploring a synthesized data set of an inventory at a large sporting goods store. The inventory pertains to different types of sports balls, 1200 in total.

---

### 3. Exploring Dataset Structure

```{r}
# Use glimpse to explore the structure of the dataset
glimpse(store_ball_inventory)
```


---

### 4. Population Parameters

```{r}
# Create a count of ball_type
store_ball_inventory |> 
  count(ball_type)

# Determine the proportion of pickleballs in the inventory
p_df <- store_ball_inventory |> 
  summarize(prop_pickle = mean(ball_type == "Pickleball"))

# Convert p to a numeric value
p <- p_df$prop_pickle

# Or using the tidyverse
p <- p_df |> pull(prop_pickle)
```


---

### 5. Take a Sample of Balls

```{r}
# Retrieve a sample of 50 balls from the inventory
ball_sample <- store_ball_inventory |> 
  slice_sample(n = 50, replace = FALSE)

# Determine the proportion of pickleballs in the sample
ball_sample |> 
  summarize(prop_pickle = mean(ball_type == "Pickleball"))
```

---

### 6. Take Another Sample of Balls

```{r}
# Retrieve another sample of 50 balls from the inventory
ball_sample2 <- store_ball_inventory |> 
  slice_sample(n = 50, replace = FALSE)

# Determine the proportion of pickleballs in the sample
ball_sample2 |> 
  summarize(prop_pickle = mean(ball_type == "Pickleball"))
```


---

### 7. Take 1000 Samples of Size 50 from the Population

```{r}
# Use `rep_slice_sample()` from the `infer` package
ball_samples <- store_ball_inventory |> 
  rep_slice_sample(n = 50, reps = 1000, replace = FALSE)
```


---

### 8. Calculate Sample Proportions and Visualize Them

```{r}
# Determine sample proportions with `dplyr`
props_pickle <- ball_samples |> 
  summarize(prop_pickle = mean(ball_type == "Pickleball"))

# Create a histogram of the sample proportions
ggplot(props_pickle, aes(x = prop_pickle)) +
  geom_histogram(bins = 15, color = "white") +
  labs(x = "Sample proportion", title = "Histogram of 1000 sample proportions of Pickleballs") 

```


---

### 9. Determine a Guess at the Standard Error

```{r}
# Using the simulations, calculate the standard deviation of the 
# sample proportions
se_sample_props <- props_pickle |> 
  summarize(sd(prop_pickle)) |> 
  pull()

# Using the formula for the standard error of a sample proportion
n <- 50
se_sample_props_formula <- sqrt(p * (1 - p) / n)
```


---

<!-- Need to be written -->

### Session 1 Review Questions

**(1.1)** What is the purpose of the `na.omit()` function in the code below?

```r
penguins_data <- penguins |>
select(species, island, flipper_length_mm, body_mass_g) |>
na.omit()
```

A. It replaces missing values with the median.  
B. It removes any rows that contain missing values.  
C. It converts missing values to zeros.  
D. It fills missing values with the previous non-missing value.

---

**(1.2)** What does the `tidy_summary()` function do when applied to numeric columns in the `penguins_data` data frame?

A. It generates summary statistics for each column in the data frame.  
B. It prints the first 6 rows of the data frame.  
C. It provides a compact overview of the data, including column names and data types.  
D. It creates a scatterplot of the variables in the data frame.

---

**(1.3)** Which of the following correctly calculates the mean body mass in the `penguins_data` data frame?

A. `summarize |> penguins_data(mean_flipper = mean(body_mass_g))`  
B. `summarize(mean(body_mass_g))`  
C. `summarize(penguins_data, mean = body_mass_g)`
D. `penguins_data |> summarize(mean_body_mass = mean(body_mass_g))`  

---

**(1.4)** What does the `get_correlation()` function do when applied in the code below?

```r
penguins_data |>
get_correlation(formula = flipper_length_mm ~ body_mass_g)
```

A. It fits a linear regression model.  
B. It gives a measure of the linear relationship between flipper length and body mass.
C. It plots a scatterplot of flipper length and body mass.  
D. It generates summary statistics for flipper length and body mass.

---

**(1.5)** In the following code, what is the purpose of `geom_smooth(method = "lm", se = FALSE)`?

```r
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g)) +
geom_point(alpha = 0.5) +
geom_smooth(method = "lm", se = FALSE) +
labs(x = "Flipper Length (mm)", y = "Body Mass (g)",
title = "Flipper Length vs. Body Mass with Regression Line")
```

A. It adds a smoothed curve based on a polynomial fit to the scatterplot.  
B. It adds a linear regression line to the scatterplot without displaying the confidence interval.  
C. It adjusts the transparency of the points in the scatterplot.  
D. It calculates and displays residuals on the plot.

---

## Session 2: Estimation using Theory-Based Methods


---

### Session 2 Review Questions


**(2.1)** What is the default baseline island in the simple linear regression model for predicting `body_mass_g` using `island` as a categorical variable?

A. Torgersen  
B. Biscoe  
C. Dream  
D. The island with the highest body mass

---

**(2.2)** In the following regression model with interaction terms, what does the coefficient for `flipper_length_mm:speciesGentoo` represent?

```r
multi_model_interaction <- lm(body_mass_g ~ flipper_length_mm * species, data = penguins_data)
```

A. The slope of the regression line for Gentoo penguins  
B. The average body mass of Gentoo penguins  
C. The change in the relationship between flipper length and body mass for Gentoo penguins compared to the baseline  
D. The difference in intercept for Gentoo penguins compared to the baseline  

---

**(2.3)** What is the purpose of adding interaction terms to a multiple regression model?

A. To make the model more complex without any real benefit  
B. To estimate the relationship between each explanatory variable and the response variable independently  
C. To account for how the relationship between one explanatory variable and the response depends on another explanatory variable  
D. To automatically improve the model's R-squared value

---

**(2.4)** What does the following plot indicate about the relationship between flipper length and body mass for different islands?

```{r}
ggplot(penguins_data, aes(x = flipper_length_mm, y = body_mass_g, color = island)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Flipper Length (mm)", y = "Body Mass (g)", color = "Island")
```

A. The relationship between flipper length and body mass is the same for all islands 
B. The relationship between flipper length and body mass differs across species, with each island having a different slope  
C. The islands do not influence the relationship between flipper length and body mass  
D. Only the intercept differs for each island, but the slope is the same

---

**(2.5)** In the multiple regression model with interaction terms, how are dummy variables used for categorical predictors?

A. Dummy variables are used to represent the different categories of a categorical variable 
B. Dummy variables are used to represent each numerical variable  
C. Dummy variables are used to replace the intercept in the regression model  
D. Dummy variables are not necessary for categorical variables

---

## Session 3: Multiple Linear Regression Analysis (Part 2)

---

### Session 3 Review Questions

**`(3.1)`** What is the main assumption in a multiple regression model without interaction terms?

A. The relationship between the explanatory variables and the response is always quadratic.  
B. The slope of the regression line is different for each level of the categorical variable.  
C. The slope of the regression line is the same for all levels of the categorical variable.  
D. The intercept of the regression line is the same for all levels of the categorical variable.

---

**`(3.2)`** In the model `lm(body_mass_g ~ flipper_length_mm + species, data = penguins_data)`, what does the coefficient for `speciesGentoo` represent?

A. The effect of being a Gentoo penguin on body mass compared to the baseline species (Adélie).  
B. The effect of flipper length on body mass for Gentoo penguins only.  
C. The change in flipper length due to species.  
D. The effect of body mass on flipper length for Gentoo penguins.

---

**`(3.3)`** What does the function `geom_parallel_slopes()` do in a regression plot?

A. It fits regression lines with different slopes for each group.  
B. It plots regression lines with the same slope but different intercepts for each group.  
C. It creates a scatterplot without any regression lines.  
D. It visualizes interaction effects between the variables.

---

**`(3.4)`** In the model `lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins_data)`, what do the coefficients for `flipper_length_mm` and `bill_length_mm` represent?

A. The total body mass of each penguin species.  
B. The predicted body mass for penguins with average bill and flipper lengths.
C. The interaction effect between flipper length and bill length.  
D. The effect of bill length on body mass, controlling for flipper length, and vice versa. 

---

**`(3.5)`** What does a regression model without interaction terms assume about the relationship between the explanatory variables and the response variable?

A. The explanatory variables are not related to the response.  
B. The relationship between the explanatory variables and the response is independent of one another.  
C. The explanatory variables interact to affect the response.  
D. The response variable depends only on the categorical variables.
