---
title: "Statistics in R with the `tidyverse`"
author: "Dr. Chester Ismay"
subtitle: "Day 2 Exercises"
format: html
---

```{r}
#| include: false
options(width = 120)
```


# Day 2: Linear Regression in R

## Session 4: Simple Linear Regression Analysis

### 1. Load Necessary Libraries

*We did this during the course in walkthroughs, but if you haven't done it yet:*

```{r}
# Load the required libraries
library(dplyr)
library(ggplot2)
library(moderndive)
library(readr)
```

- These packages provide tools for data wrangling, visualization, and modeling.  

---

### 2. Prepare Dataset

```{r}
# Load in the data_dev_survey CSV file as a data frame again
# selecting the columns of interest keeping only complete rows of information
data_dev_survey <- read_csv("../data_dev_survey.csv") |> 
  select(response_id, converted_comp_yearly, work_exp, 
         years_code, ai_trust) |> 
  # This code sets the levels of the ai_trust column to be more intuitive
  # Note this will change the baseline level for the factor!
  mutate(ai_trust = factor(ai_trust, levels = c("Highly distrust", 
                                                "Somewhat distrust",
                                                "Neither trust nor distrust",
                                                "Somewhat trust",
                                                "Highly trust"))) |> 
  na.omit()
```

- We're still exploring the `data_dev_survey` data from Day 1 here.

---

### 3. Exploring Dataset Structure

```{r}
# Use glimpse to remind yourself of the structure of the data

```

---

### 4. Summary Statistics

```{r}
# Summarize mean and median values for converted_comp_yearly and work_exp

```

---

### 5. Using `tidy_summary()` Function

```{r}
# Use tidy_summary to get a clean summary of the two columns

```

---

### 6. Correlation Between Salary and Work Experience

```{r}
# Compute the correlation between converted_comp_yearly and work_exp

```

---

### 7. Scatterplot of Salary and Work Experience

```{r}
# Plot the relationship between work experience and yearly compensation

```

---

### 8. Regression Line

```{r}
# The same scatterplot with a regression line added

```

---

### 9. Fit a Regression Model

```{r}
# Fit a linear regression model with compensation as the outcome


# Get the regression coefficients

```

---

## Session 5: Multiple Linear Regression Analysis (Part 1)

### 10: Fitting a Simple Linear Regression with a Categorical Regressor

```{r}
# Fit a simple linear regression model with ai_trust as a predictor on salary


# Get regression coefficients

```

---

### 11: Visualizing Data with Categorical Variable

```{r}
# Scatterplot with ai_trust as color for grouping on work_exp predicting
# converted_comp_yearly

```

---

### 12: Summarizing the Target and Regressors (One numerical and one categorical)

```{r}
# Summarize the target and regressors

```

---

### 13: Fitting a Multiple Regression Model with Interaction Terms

```{r}
# Fit a multiple regression model with interaction terms between
# work experience and trust in AI 


# Get regression coefficients


```

---

### 14: Adding Interaction Terms to the Regression Plot

```{r}
# Scatterplot with regression lines by trust in AI (interaction terms)

```

---

## Session 6: Multiple Linear Regression Analysis (Part 2)

### 15. Fitting a Multiple Regression Model Without Interactions

```{r}
# Fit a multiple regression model without interaction terms


# Get regression coefficients

```

---

### 16. Visualizing the Parallel Slopes Model

```{r}
# Scatterplot with regression lines by trust in AI (parallel slopes)

```

---

### 17. Summarize the Numerical Regressors and Target Variable

```{r}
# Use tidy_summary to get a clean summary of the data for a multiple regression
# with work experience and years coding as regressors and yearly compensation as
# the outcome

```

---

### 18. Fitting a Multiple Regression Model with Two Numerical Predictors

```{r}
# Fit a multiple regression model


# Get regression coefficients

```

---

### 19. Visualizing the Relationship between Variables (with SLR)

```{r}
# Create scatterplots with regression lines for both variables with 
# converted_comp_yearly as the response variable in simple linear regression


```

